Ecosystem size and complexity dictate riverine biodiversity

### 2021-03-17 manuscript_sciad.Rmd
edits: maintext, minor

However, dispersal processes affected mechanisms that underlie the positive effects of ecosystem size and complexity on $\gamma$ diversity (compare left and right columns in **Figures 2 and 3**). We observed a greater contribution of $\beta$ diversity (defined as $\frac{\gamma}{\alpha}$) to increased $\gamma$ diversity when dispersal limitation existed (i.e., species travel short distances). This pattern reflected significant spatial variation in species composition over the branching network and was likely driven by the local association of species' niche and abiotic environments (i.e., species sorting) [@leiboldMetacommunityConceptFramework2004; @thompsonProcessbasedMetacommunityFramework2020]. In contrast, when the dispersal limitation was relaxed (species travel long distances), a clear increase in $\alpha$ diversity underpinned the positive relationships between $\gamma$ diversity and ecosystem properties. The results agree with previous predictions that increased dispersal homogenizes community composition while enhancing local diversity through increased immigrants from suitable habitat patches (i.e., mass effects) [@leiboldMetacommunityConceptFramework2004; @economoSpeciesDiversityNeutral2008; @thompsonProcessbasedMetacommunityFramework2020]. Therefore, ecosystem size and complexity can increase $\gamma$ diversity via enhanced species sorting or mass effects, and which takes primacy is dependent on the degree of dispersal limitation. These patterns were consistent across different levels of dispersal probabilities (**Figures S5 and S12**). The strength of competition decreased the maximum of $\alpha$ diversity but did not change the scaling relationships with ecosystem properties (**Figures 2 and 3**). In summary, our theory highlights how apparently similar patterns in $\gamma$ diversity emerge through different ecological pathways.

@@

We found decisive support for the global model in explaining $\gamma$ diversity (Bayes factor = `r round(fit$gamma$bayes_factor, 1)`) despite the substantial difference in fish fauna between the study regions (**Tables S5 and S6**). The estimated $\gamma$ diversity linearly increased with increasing watershed area (ecosystem size) and branching probability (ecosystem complexity) on a log-log scale (i.e., power-law), and these patterns were consistent across the study regions (**Table 1 and Figure 4**). Thus, watershed-scale riverine biodiversity likely obeys the general scaling law along the two orthogonal dimensions of ecosystem structure, i,e, size and complexity. In particular, the effect of ecosystem complexity was striking in its magnitude. Average predictive comparisons [@gelmanAveragePredictiveComparisons2007] revealed that an expected increase of $\gamma$ diversity per 0.1 branching probability was `r op(ap_est$sim_estimate[2] * 0.1)` $\pm$ `r op(ap_est$sim_se[2] * 0.1)` species (**Materials and Methods** and **Supplementary Materials**); this level of increase in $\gamma$ diversity requires `r round(ap_est$sim_estimate[2] * 0.1 / ap_est$sim_estimate[1])` km^2^ in the watershed area (see **Table S7** for the estimated average predictive comparisons). This result is reasonable because numerous studies have shown that tributaries are the primary source of environmental heterogeneity in river networks [@brownDendriticNetworkStructure2010; @petersonModellingDendriticEcological2013; @mooreEmergentStabilityLarge2015; @teruiMetapopulationStabilityBranching2018; @isaakSlowClimateVelocities2016; @mcguireNetworkAnalysisReveals2014; @abbottUnexpectedSpatialStability2018] - the sole theoretical assumption required for the dual scaling of $\gamma$ diversity to emerge. Further, the consistent patterns across the geographic regions with distinct evolutionary histories imply that ecological traits of species in metacommunities may have a minor effect on the scaling relationships, as predicted by our theory.

>>>>

However, dispersal processes affected mechanisms that underlie the positive effects of ecosystem size and complexity on $\gamma$ diversity (compare left and right columns in **Figures 2 and 3**). We observed a greater contribution of $\beta$ diversity (defined as $\frac{\gamma}{\alpha}$) to increased $\gamma$ diversity when dispersal limitation existed (i.e., species travel short distances). This pattern reflected significant spatial variation in species composition over the branching network and was likely driven by the local association of species' niche and abiotic environments (i.e., species sorting) [@leiboldMetacommunityConceptFramework2004; @thompsonProcessbasedMetacommunityFramework2020]. In contrast, when the dispersal limitation was relaxed (species travel long distances), a clear increase in $\alpha$ diversity underpinned the positive relationships between $\gamma$ diversity and ecosystem properties. The results agree with previous predictions that increased dispersal homogenizes community composition while enhancing local diversity through increased immigrants from suitable habitat patches (i.e., mass effects) [@leiboldMetacommunityConceptFramework2004; @economoSpeciesDiversityNeutral2008; @thompsonProcessbasedMetacommunityFramework2020]. Therefore, ecosystem size and complexity can increase $\gamma$ diversity via enhanced species sorting or mass effects, and which takes primacy is dependent on the degree of dispersal limitation. These patterns were qualitatively similar across different levels of dispersal probabilities (**Figures S5 and S12**). The strength of competition decreased the maximum of $\alpha$ diversity but did not change the scaling relationships with ecosystem properties (**Figures 2 and 3**). In summary, our theory highlights how apparently similar patterns in $\gamma$ diversity emerge through different ecological pathways.

@@

We found decisive support for the global model in explaining $\gamma$ diversity (Bayes factor = `r round(fit$gamma$bayes_factor, 1)`) despite the substantial difference in fish fauna between the study regions (**Tables S5 and S6**). The estimated $\gamma$ diversity linearly increased with increasing watershed area (ecosystem size) and branching probability (ecosystem complexity) on a log-log scale (i.e., power-law), and these patterns were consistent across the study regions (**Table 1 and Figure 4**). Thus, watershed-scale riverine biodiversity likely obeys a general scaling law along the two orthogonal dimensions of ecosystem structure, i,e, size and complexity. In particular, the effect of ecosystem complexity was striking in its magnitude. Average predictive comparisons [@gelmanAveragePredictiveComparisons2007] revealed that an expected increase of $\gamma$ diversity per 0.1 branching probability was `r op(ap_est$sim_estimate[2] * 0.1)` $\pm$ `r op(ap_est$sim_se[2] * 0.1)` species (**Materials and Methods** and **Supplementary Materials**); this level of increase in $\gamma$ diversity requires `r round(ap_est$sim_estimate[2] * 0.1 / ap_est$sim_estimate[1])` km^2^ in the watershed area (see **Table S7** for the estimated average predictive comparisons). This result is reasonable because numerous studies have shown that tributaries are the primary source of environmental heterogeneity in river networks [@brownDendriticNetworkStructure2010; @petersonModellingDendriticEcological2013; @mooreEmergentStabilityLarge2015; @teruiMetapopulationStabilityBranching2018; @isaakSlowClimateVelocities2016; @mcguireNetworkAnalysisReveals2014; @abbottUnexpectedSpatialStability2018] - the sole theoretical assumption required for the dual scaling of $\gamma$ diversity to emerge. Further, the consistent patterns across the geographic regions with distinct evolutionary histories imply that ecological traits of species in metacommunities may have a minor effect on the scaling relationships, as predicted by our theory.
